---
title: "KS Tests"
format: html
execute:
	warning: false
	message: false
---

```{r}
library(knitr)

find_repo_root <- function() {
	candidates <- normalizePath(
		c(getwd(), file.path(getwd(), ".."), file.path(getwd(), "..", "..")),
		winslash = "/",
		mustWork = FALSE
	)
	candidates <- unique(candidates[file.exists(file.path(candidates, "R"))])
	candidates <- candidates[file.exists(file.path(candidates, "results"))]
	if (length(candidates) == 0) {
		stop("Could not locate the repository root with both R/ and results/ directories.")
	}
	candidates[1]
}

repo_root <- find_repo_root()
results_root <- file.path(repo_root, "results")
batch_dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
batch_dirs <- batch_dirs[grepl("^noise_k_batch_", basename(batch_dirs))]
if (length(batch_dirs) == 0) {
	stop("No noise_k batch results were found under results/.")
}

batch_info <- data.frame(
	dir = batch_dirs,
	mtime = file.info(batch_dirs)$mtime,
	stringsAsFactors = FALSE
)
latest_dir <- batch_info$dir[which.max(batch_info$mtime)]
summary_path <- file.path(latest_dir, "noise_k_summary.csv")
curve_path <- file.path(latest_dir, "noise_k_all_curves.csv")

if (!file.exists(summary_path)) {
	stop(paste("Missing summary CSV:", summary_path))
}

summary_results <- read.csv(summary_path, stringsAsFactors = FALSE)
curve_results <- if (file.exists(curve_path)) read.csv(curve_path, stringsAsFactors = FALSE) else data.frame()

summary_results$size_label <- paste0(summary_results$nrow_matrix, " x ", summary_results$ncol_matrix)
summary_results$status <- factor(summary_results$status)
ok_results <- summary_results[summary_results$status == "ok", , drop = FALSE]

display_floor <- 10^-4
ok_results$is_exact_zero_k <- ok_results$selected_k == 0
ok_results$is_tiny_positive_k <- ok_results$selected_k > 0 & ok_results$selected_k < display_floor

size_levels <- unique(summary_results$size_label)
size_levels <- size_levels[order(match(size_levels, unique(summary_results$size_label)))]
ok_results$size_label <- factor(ok_results$size_label, levels = size_levels)

summary_table <- do.call(
	rbind,
	lapply(split(ok_results, ok_results$size_label), function(df) {
		data.frame(
			size = unique(df$size_label),
			n_runs = nrow(df),
			contam_min = min(df$n_contam),
			contam_max = max(df$n_contam),
			selected_k_mean = mean(df$selected_k, na.rm = TRUE),
			selected_k_median = median(df$selected_k, na.rm = TRUE),
			ks_mean = mean(df$selected_ks, na.rm = TRUE),
			ks_median = median(df$selected_ks, na.rm = TRUE),
			stringsAsFactors = FALSE
		)
	})
)

failure_table <- do.call(
	rbind,
	lapply(split(summary_results, summary_results$size_label), function(df) {
		data.frame(
			size = unique(df$size_label),
			n_runs = nrow(df),
			failures = sum(df$status != "ok"),
			failure_rate = mean(df$status != "ok"),
			stringsAsFactors = FALSE
		)
	})
)

contam_table <- aggregate(
	cbind(selected_k, selected_ks) ~ size_label + n_contam,
	data = ok_results,
	FUN = function(x) c(mean = mean(x, na.rm = TRUE), median = median(x, na.rm = TRUE))
)

selected_k_by_contam <- aggregate(selected_k ~ size_label + n_contam, data = ok_results, FUN = median)
selected_ks_by_contam <- aggregate(selected_ks ~ size_label + n_contam, data = ok_results, FUN = median)

k_diagnostic_table <- do.call(
	rbind,
	lapply(split(ok_results, ok_results$size_label), function(df) {
		positive_k <- df$selected_k[df$selected_k > 0]
		min_positive_k <- if (length(positive_k) > 0) min(positive_k) else NA_real_
		data.frame(
			size = unique(df$size_label),
			n_runs = nrow(df),
			n_exact_zero = sum(df$is_exact_zero_k, na.rm = TRUE),
			n_tiny_positive = sum(df$is_tiny_positive_k, na.rm = TRUE),
			min_positive_k = min_positive_k,
			min_positive_k_sci = if (is.finite(min_positive_k)) format(min_positive_k, scientific = TRUE, digits = 6) else NA_character_,
			stringsAsFactors = FALSE
		)
	})
)

global_k_check <- data.frame(
	n_total_ok = nrow(ok_results),
	n_exact_zero = sum(ok_results$is_exact_zero_k, na.rm = TRUE),
	n_tiny_positive = sum(ok_results$is_tiny_positive_k, na.rm = TRUE),
	stringsAsFactors = FALSE
)

example_run <- ok_results[which.min(ok_results$selected_ks), , drop = FALSE]
example_curve <- if (nrow(example_run) == 1 && file.exists(example_run$refined_curve_file)) {
	read.csv(example_run$refined_curve_file, stringsAsFactors = FALSE)
} else {
	data.frame()
}

if (nrow(example_curve) > 0) {
	example_curve$selected <- FALSE
	example_curve$selected[which.min(example_curve$ks_statistic)] <- TRUE
}
```

## Latest Batch

The report reads the most recent batch output under `results/noise_k_batch_*`.

```{r}
cat("Latest results folder:", latest_dir, "\n")
cat("Rows in summary:", nrow(summary_results), "\n")
cat("Successful runs:", nrow(ok_results), "\n")
cat("Failed runs:", sum(summary_results$status != "ok"), "\n")
```

## Summary Tables

### Selected k and KS by matrix size

```{r}
kable(summary_table, digits = 4)
```

### Failure rate by matrix size

```{r}
kable(failure_table, digits = 4)
```

### Selected k zero-vs-printing diagnostic

This checks whether `selected_k` is exactly zero, or just very small positive and rounded to zero in printed tables.

```{r}
kable(global_k_check)
kable(k_diagnostic_table, digits = 6)
```

## Selected k by Contamination

```{r}
op <- par(mar = c(4.5, 4.5, 3.5, 1), mfrow = c(1, 1))
on.exit(par(op), add = TRUE)

plot(
	NA,
	xlim = range(selected_k_by_contam$n_contam),
	ylim = range(log10(selected_k_by_contam$selected_k)),
	xlab = "Contamination count",
	ylab = expression(log[10](selected[k])),
	main = "Median selected k by contamination",
)

cols <- seq_along(size_levels)
for (i in seq_along(size_levels)) {
	size_i <- size_levels[i]
	df_i <- selected_k_by_contam[selected_k_by_contam$size_label == size_i, , drop = FALSE]
	if (nrow(df_i) == 0) next
	lines(df_i$n_contam, log10(df_i$selected_k), col = cols[i], lwd = 2, type = "b", pch = 19)
}
#legend("topleft", legend = size_levels, col = cols, lty = 1, pch = 19, bty = "n")
```

## KS Statistic by Contamination

```{r}
plot(
	NA,
	xlim = range(selected_ks_by_contam$n_contam),
	ylim = range(selected_ks_by_contam$selected_ks, na.rm = TRUE),
	xlab = "Contamination count",
	ylab = "Median KS statistic",
	main = "Median KS statistic by contamination"
)

for (i in seq_along(size_levels)) {
	size_i <- size_levels[i]
	df_i <- selected_ks_by_contam[selected_ks_by_contam$size_label == size_i, , drop = FALSE]
	if (nrow(df_i) == 0) next
	lines(df_i$n_contam, df_i$selected_ks, col = cols[i], lwd = 2, type = "b", pch = 19)
}
#legend("topleft", legend = size_levels, col = cols, lty = 1, pch = 19, bty = "n")
```

