---
title: "Matrix_Full"
format: pdf
---

# General Question
We cluster the trends of measurements across time and group subjects with similar patterns across rows and columns.
U row covariance asks do measurements at occasions change together? (Across time)
V col covariance asks do measurements tend to change together? (Time flat)

# Matrix-Variate Normal Distribution

- n is the number of subjects
- r is the number of occasions (time points)
- p is the number of attributes

$$
\phi_{r\times p}(\mathbf{X} | \mathbf{M}, \mathbf{V}, \mathbf{U}) = \frac{1}{(2\pi)^{\frac{rp}{2}}|\mathbf{V}|^{\frac{r}{2}}|\mathbf{U}|^{\frac{p}{2}}} \exp \left\{-\frac{1}{2}\text{tr}\left(\mathbf{V}^{-1}(\mathbf{X}-\mathbf{M})^\top\mathbf{U}^{-1}(\mathbf{X}-\mathbf{M})\right) \right\}
$$

- M is a r x p mean matrix
- U is a r x r column covariance matrix
- V is a p x p column row covariance matrix

Note 1: that there is an identifiability issue with regard to U and V. If $k$ is a strictly positive constant, that replacing U and V by $(1/k)\mathbf{U}$ and $k\mathbf{V}$ respectively, leaves the distribution unchanged. Solutions have been proposed including setting $\text{tr}(\mathbf{U}) = r$ or $\mathbf{U}_{11} = 1$.

$$
\frac{1}{k}\mathbf{V} \otimes k\mathbf{U} = \mathbf{V} \otimes \mathbf{U}, \qquad \text{if} \ k >0
$$

Note 2: that the matrix-variate normal distribution is related to the multivariate normal distribution via the equivalence

$$
\mathbf{X} \sim \mathcal{N}_{r\times p} (\mathbf{M}, \mathbf{U}, \mathbf{V}) \iff \text{vec}({\mathbf{X}}) \sim \mathcal{N}_{rp}(\text{vec}(\mathbf{M}), \mathbf{V} \otimes \mathbf{U})
$$

The log-likelihood is defined as
$$
-\frac{1}{2}(rp\log(2\pi) + r\log|\mathbf{V}|+p\log|\mathbf{U}| + \text{tr}[\mathbf{V}^{-1}(\mathbf{X}-\mathbf{M})^\top\mathbf{U}^{-1}(\mathbf{X}-\mathbf{M})])
$$

## EM Updates

Let $z_{ig} = 1$ when observation $i$ belongs in group $g$, and $z_{ig} = 0$ else. Then, in the E-step we estimate the posterior $\hat{z}_{ig}$ that observation $\mathbf{X}_i$ belongs to group $g$

$$
\hat{z}_{ig} = \frac{\hat\pi_g\phi_{r\times p}\left(\mathbf{X}_i|\hat{\mathbf{M}}_g,\hat{\mathbf{V}}_g, \hat{\mathbf{U}}_g\right)}{\sum_{h=1}^G\pi_h\phi_{r\times p}\left(\mathbf{X}_i|\hat{\mathbf{M}}_h, \hat{\mathbf{V}}_h, \hat{\mathbf{U}}_h\right)}
$$

In M-step we get the maximum likelihood estimates:

$$
\begin{align}

\hat\pi_g &= \frac{1}{n}\sum_{i=1}^n\hat{z}_{ig} \\

\hat{\mathbf{M}}_g &= \frac{1}{\sum_{i=1}^n\hat{z}_{ig}}\sum_{i=1}^n\hat{z}_{ig}\mathbf{X}_i \\

\hat{\mathbf{U}}_g &= \frac{1}{p\sum_{i=1}^n\hat{z}_{ig}}\sum_{i=1}^n\hat{z}_{ig}(\mathbf{X}_i - \hat{\mathbf{M}}_g)\hat{\mathbf{V}}_g^{-1}(\mathbf{X}_i - \hat{\mathbf{M}}_g)^\top \\

\hat{\mathbf{V}}_g &= \frac{1}{r\sum_{i=1}^n\hat{z}_{ig}}\sum_{i=1}^n\hat{z}_{ig}(\mathbf{X}_i - \hat{\mathbf{M}}_g)^\top\hat{\mathbf{U}}_g^{-1}(\mathbf{X}_i - \hat{\mathbf{M}}_g)

\end{align}
$$

We can initialize $\hat{z}_{ig}$ using k-means by vectorizing the matrices first. 

Meaning we define the Mahalonobis distance as

$$
\mathcal{D}_M(\mathbf{X}_i, \mathbf{M}, \mathbf{V}, \mathbf{U}) = \text{tr}\left\{\mathbf{U}^{-1}(\mathbf{X}_i - \mathbf{M})\mathbf{V^{-1}(\mathbf{X}_i - \mathbf{M})^\top} \right\}
$$

These will form as the foundation for our base Matrix Gaussian Mixture Model (MGMM) updates. Let us try putting that into a function!

## MGMM Implementation

```{r}
#| label: libraries
#| message: false
#| warning: false
#| include: false

library(dplyr)
library(clusterGeneration)
library(ggplot2)
```

- `matrix_mahalanobis()` computes the matrix-variate Mahalanobis distance from one matrix to a component mean under the row and column covariance structure. It returns tr(U^{-1} (X - M) V^{-1} (X - M)^T).

- `matrix_mixture_kmeans_init()` builds the starting values for the mixture model using k-means on vectorized matrices. It returns initial mixing proportions, component mean matrices, row covariances, column covariances, and the starting cluster labels.

- `make_spd` forces a non positive-definite matrix to be positive definite. It makes the matrix symmetric by averaging it with its transpose, and if the cholesky decomp fails it adds a bit of noise and retrys the decomposition until it works. 

```{r}
#| label: Mahalonbis_D
#| warning: false
#| message: false

# Force a matrix to be symmetric positive definite for stable chol/solve
# Not having this caused issues during the viroli and tomarchio sims
make_spd <- function(mat, jitter = 1e-8, max_tries = 8) {
	mat <- (mat + t(mat)) / 2
	for (k in 0:max_tries) {
		j <- jitter * (10^k)
		candidate <- mat + diag(j, nrow(mat))
		ok <- tryCatch({
			chol(candidate)
			TRUE
		}, error = function(e) FALSE)
		if (ok) return(candidate)
	}
	stop("Could not make covariance matrix positive definite.")
}

# Compute matrix-variate Mahalanobis distance
matrix_mahalanobis <- function(x, mean_matrix, row_cov, col_cov) {
	# U^{-1} and V^{-1}
	row_cov <- make_spd(row_cov)
	col_cov <- make_spd(col_cov)
	row_chol <- chol(row_cov)
	col_chol <- chol(col_cov)
	centered <- x - mean_matrix
	row_inv_centered <- backsolve(row_chol, forwardsolve(t(row_chol), centered))
	col_inv <- chol2inv(col_chol)

	sum(row_inv_centered * (centered %*% col_inv))
}

# kmeans initialization
matrix_mixture_kmeans_init <- function(x_list, g, nstart = 10) {
	# Extract dimensions
	n <- length(x_list)
	r <- nrow(x_list[[1]])
	p <- ncol(x_list[[1]])

	# vectorize and run kmeans for init
	x_matrix <- do.call(rbind, lapply(x_list, function(x) as.vector(x)))
	km <- kmeans(x_matrix, centers = g, nstart = nstart)
	z <- km$cluster

	mixing_proportions <- numeric(g)
	mean_matrices <- vector("list", g)
	row_covariances <- vector("list", g)
	col_covariances <- vector("list", g)

	# For each component, compute sample mean and covariances from k-means clusters
	for (component in seq_len(g)) {
		component_index <- which(z == component)
		if (length(component_index) == 0) {
			component_index <- sample.int(n, 1)  # Fallback if k-means created empty cluster
		}

		component_data <- x_list[component_index]
		mixing_proportions[component] <- length(component_index) / n
		# Calculates M_hat
		mean_matrices[[component]] <- Reduce(`+`, component_data) / length(component_data)

		# M-step for the covariance U and V
		row_cov <- matrix(0, r, r)
		col_cov <- matrix(0, p, p)
		for (x in component_data) {
			centered <- x - mean_matrices[[component]]
			row_cov <- row_cov + centered %*% t(centered)
			col_cov <- col_cov + t(centered) %*% centered
		}
		
		row_cov <- row_cov / (p * length(component_data))
		col_cov <- col_cov / (r * length(component_data))
		row_cov <- make_spd(row_cov)
		col_cov <- make_spd(col_cov)

		# Enforce identifiability scale so that tr(U) = r
		row_covariances[[component]] <- row_cov
		col_covariances[[component]] <- col_cov
		row_scale <- r / sum(diag(row_covariances[[component]]))
		row_covariances[[component]] <- row_covariances[[component]] * row_scale
		col_covariances[[component]] <- col_covariances[[component]] / row_scale
		row_covariances[[component]] <- make_spd(row_covariances[[component]])
		col_covariances[[component]] <- make_spd(col_covariances[[component]])
	}

	list(
		pi = mixing_proportions,
		M = mean_matrices,
		U = row_covariances,
		V = col_covariances,
		cluster = z
	)
}
```

- `matrix_variate_log_density()` (Z_igs posterior E-step Update) computes the log-density of a single matrix under the matrix normal distribution. It uses the row covariance and column covariance separately.

- `matrix_variate_mixture_fit()` is the main EM fitter for the matrix variate Gaussian mixture model. It repeatedly does the E-step to compute responsibilities and the M-step to update mixing proportions, means, row covariances, and column covariances until convergence or the iteration limit is reached.

```{r}
#| label: MGMM_base
#| warning: false
#| message: false

# Compute log-density of single matrix under matrix-variate normal
matrix_variate_log_density <- function(x, mean_matrix, row_cov, col_cov) {
	# Cholesky decomposition
	row_cov <- make_spd(row_cov)
	col_cov <- make_spd(col_cov)
	row_chol <- chol(row_cov)
	col_chol <- chol(col_cov)

	# |U| and |V| for the denominator
	row_logdet <- 2 * sum(log(diag(row_chol)))
	col_logdet <- 2 * sum(log(diag(col_chol)))

	# tr(V^{-1} * (X - M)^T * U^{-1} * (X - M))
	centered <- x - mean_matrix
	row_inv_centered <- backsolve(row_chol, forwardsolve(t(row_chol), centered))
	col_inv <- chol2inv(col_chol)
	trace_form <- sum(row_inv_centered * (centered %*% col_inv))

	r <- nrow(x)
	p <- ncol(x)

	# Returns log density value
	-0.5 * (r * p * log(2 * pi) + p * row_logdet + r * col_logdet + trace_form)
}

matrix_variate_mixture_fit <- function(x_list, g, max_iter = 100, tol = 1e-06,
																			 nstart = 10, verbose = FALSE) {
	if (!is.list(x_list) || length(x_list) == 0) {
		stop("x_list must be a non-empty list of matrices.")
	}

	n <- length(x_list)
	r <- nrow(x_list[[1]])
	p <- ncol(x_list[[1]])

	for (x in x_list) {
		if (!is.matrix(x) || nrow(x) != r || ncol(x) != p) {
			stop("All elements of x_list must be matrices with the same dimensions.")
		}
	}

	# Initialize parameters using k-means
	params <- matrix_mixture_kmeans_init(x_list, g = g, nstart = nstart)
	loglik_trace <- numeric(0)
	responsibilities <- matrix(0, n, g)  # Will hold posterior probabilities P(z_ig = 1 | X_i)

	log_sum_exp <- function(values) {
		max_value <- max(values)
		max_value + log(sum(exp(values - max_value)))
	}

	# EM loop
	for (iteration in seq_len(max_iter)) {
		# E-step
		log_density <- matrix(NA_real_, nrow = n, ncol = g)

		for (component in seq_len(g)) {
			for (i in seq_len(n)) {
				# compute P(X_i | component)
				log_density[i, component] <- log(params$pi[component]) +
					matrix_variate_log_density(
						x = x_list[[i]],
						mean_matrix = params$M[[component]],
						row_cov = params$U[[component]],
						col_cov = params$V[[component]]
					)
			}
		}

		# Normalize log-densities to get responsibilities
		for (i in seq_len(n)) {
			row_log_densities <- log_density[i, ]
			normalizer <- log_sum_exp(row_log_densities)  # numerically stable log-sum-exp
			responsibilities[i, ] <- exp(row_log_densities - normalizer)  # z_hat_ig
		}

		# Compute observed data log-likelihood for convergence check
		current_loglik <- sum(apply(log_density, 1, log_sum_exp))
		loglik_trace <- c(loglik_trace, current_loglik)

		# Check convergence
		if (iteration > 1 && abs(loglik_trace[iteration] - loglik_trace[iteration - 1]) < tol) {
			break
		}

		# M-step
		component_sizes <- colSums(responsibilities)  # sum of z_hat_ig over all observations i
		new_params <- params

		# Update each component's parameters
		for (component in seq_len(g)) {
			if (component_sizes[component] <= 0) {
				next
			}

			weights <- responsibilities[, component]
			weights_sum <- component_sizes[component]  # effective sample size for this component
			v_for_row <- make_spd(params$V[[component]])

			# Update mean matrix (M-step: M_hat_g)
			mean_matrix <- matrix(0, r, p)
			for (i in seq_len(n)) {
				mean_matrix <- mean_matrix + weights[i] * x_list[[i]]
			}
			mean_matrix <- mean_matrix / weights_sum

			# Update row covariance U_hat_g
			row_cov <- matrix(0, r, r)
			for (i in seq_len(n)) {
				centered <- x_list[[i]] - mean_matrix
				row_cov <- row_cov + weights[i] * (centered %*% solve(v_for_row, t(centered)))
			}
			row_cov <- row_cov / (p * weights_sum)
			row_cov <- make_spd(row_cov)
			
			# Enforce tr(U) = r
			row_scale <- r / sum(diag(row_cov))
			row_cov <- row_cov * row_scale
			row_cov <- make_spd(row_cov)

			# Update column covariance V_hat_g
			col_cov <- matrix(0, p, p)
			for (i in seq_len(n)) {
				centered <- x_list[[i]] - mean_matrix
				col_cov <- col_cov + weights[i] * (t(centered) %*% solve(row_cov, centered))
			}
			col_cov <- col_cov / (r * weights_sum)
			col_cov <- make_spd(col_cov)

			# Store updated parameters
			new_params$pi[component] <- weights_sum / n  # mixing proportion
			new_params$M[[component]] <- mean_matrix
			new_params$U[[component]] <- row_cov
			new_params$V[[component]] <- col_cov
		}

		# Normalize mixing proportions to sum to 1
		new_params$pi <- new_params$pi / sum(new_params$pi)

		params <- new_params

		if (verbose) {
			message(sprintf("Iteration %d: log-likelihood = %.4f", iteration, current_loglik))
		}
	}

	# Assign each observation to its most likely component
	cluster_membership <- max.col(responsibilities, ties.method = "first")

	# Return fitted model with all parameters and diagnostics
	list(
		pi = params$pi,
		M = params$M,
		U = params$U,
		V = params$V,
		z = responsibilities,
		cluster = cluster_membership,
		logLik = loglik_trace,
		iterations = length(loglik_trace),
		converged = length(loglik_trace) < max_iter
	)
}

```

## Testing the Model

```{r}
#| label: fit_ex
#| warning: false
#| message: false

set.seed(123)

simulate_matrix_group <- function(n, mean_matrix, row_sd = 0.35, col_sd = 0.35) {
	r <- nrow(mean_matrix)
	p <- ncol(mean_matrix)

	row_cov <- diag(row_sd, r)
	col_cov <- diag(col_sd, p)

	lapply(seq_len(n), function(i) {
		noise <- matrix(rnorm(r * p), r, p)
		mean_matrix + row_cov %*% noise %*% col_cov
	})
}

mean_1 <- matrix(c(1.5, 1.2, 1.0, 1.3,
					 1.1, 1.4, 1.2, 1.0), nrow = 2, byrow = TRUE)
mean_2 <- matrix(c(-1.4, -1.0, -1.2, -1.3,
					 -1.1, -1.5, -1.0, -1.2), nrow = 2, byrow = TRUE)

x_list <- c(
	simulate_matrix_group(n = 15, mean_matrix = mean_1),
	simulate_matrix_group(n = 15, mean_matrix = mean_2)
)

fit <- matrix_variate_mixture_fit(x_list = x_list, g = 2, max_iter = 50, verbose = TRUE)

fit$cluster
fit$pi
fit$logLik

true_labels <- rep(1:2, each = 15)
pred_labels <- fit$cluster

accuracy_1 <- mean(pred_labels == true_labels)
accuracy_2 <- mean((3 - pred_labels) == true_labels)
best_accuracy <- max(accuracy_1, accuracy_2)
aligned_labels <- if (accuracy_2 > accuracy_1) 3 - pred_labels else pred_labels

best_accuracy
table(True = true_labels, Predicted = aligned_labels)
```

## In the Style of Viroli

A total of 100 datasets are created with n = 300 $3 \times 5$ matrices in three groups with proportions $\pi = (0.3, 0.4, 0.3)$ and means of the form

$$
M_g = \begin{pmatrix}
      a & 0 & 0 & 0 & 0\\
      b & 0 & 0 & 0 & 0\\
      0 & 0 & 0 & 0 & 0
      \end{pmatrix}
$$

with M_1 : $(a,b)$ = $(0.5, 0.5)$, M_2 : $(a,b)$ = $(0,0)$ and M_3 : $(a,b)$ = $(-0.5, 0.5)$. U_g and V_g matrices of dimensions $3\times 3 \ \text{and} \ 5\times 5$ are randomly generated using the rcorrmatrix function in the clusterGeneration package.

```{r}
#| label: viroli_simulation_setup
#| warning: false
#| message: false

generate_viroli_means <- function() {
	# M1: (a, b) = (0.5, 0.5)
	M1 <- matrix(0, nrow = 3, ncol = 5)
	M1[1, 1] <- 0.5  # a
	M1[2, 1] <- 0.5  # b
	
	# M2: (a, b) = (0, 0)
	M2 <- matrix(0, nrow = 3, ncol = 5)
	
	# M3: (a, b) = (-0.5, 0.5)
	M3 <- matrix(0, nrow = 3, ncol = 5)
	M3[1, 1] <- -0.5  # a
	M3[2, 1] <- 0.5   # b
	
	list(M1 = M1, M2 = M2, M3 = M3)
}

generate_random_corrmatrix <- function(p, seed = NULL) {
	make_spd(rcorrmatrix(p))
}

# Generate a single Viroli dataset
generate_viroli_dataset <- function(n = 300, means_list, U_list, V_list, 
                                    group_props = c(0.3, 0.4, 0.3),
                                    seed = NULL) {
	# Generate group assignments
	group_sizes <- as.numeric(table(sample(rep(1:3, times = round(n * group_props)),
	                                        size = n)))
	groups <- rep(1:3, times = group_sizes)
	
	# List to store all matrices and their group labels
	X_list <- list()
	group_labels <- numeric(n)
	
	idx <- 1
	for (g in 1:3) {
		n_g <- group_sizes[g]
		U_g <- make_spd(U_list[[g]])
		V_g <- make_spd(V_list[[g]])
		chol_U <- chol(U_g)
		chol_V <- chol(V_g)
		for (i in 1:n_g) {
			# Generate random noise from matrix variate normal
			Z <- matrix(rnorm(3 * 5), nrow = 3, ncol = 5)
			
			# X_ig ~ N_{3x5}(M_g, U_g, V_g)
			# Equivalent to: X_ig = M_g + U_g^{1/2} * Z * V_g^{1/2}
			X_matrix <- means_list[[g]] + t(chol_U) %*% Z %*% chol_V
			X_list[[idx]] <- X_matrix
			group_labels[idx] <- g
			idx <- idx + 1
		}
	}
	
	list(X = X_list, true_groups = group_labels, group_sizes = group_sizes)
}

# Viroli simulation with N_sim datasets
run_viroli_simulation <- function(N_sim = 100, n = 300, g = 3, 
                                  verbose = TRUE) {
	
	# Generate the mean matrices
	means_list <- generate_viroli_means()
	means_components <- list(means_list$M1, means_list$M2, means_list$M3)
	
	# Covariance matrices
	U_list <- lapply(1:3, function(g) generate_random_corrmatrix(3))
	V_list <- lapply(1:3, function(g) generate_random_corrmatrix(5))
	
	# Lists to store results from each dataset
	all_accuracies <- numeric(N_sim)
	all_confusion_matrices <- list()
	all_fits <- list()
	
	# Run simulation
	for (sim in 1:N_sim) {
		if (verbose && sim %% 10 == 0) {
			cat(sprintf("Simulation %d/%d\n", sim, N_sim))
		}
		
		# Generate dataset
		dataset <- generate_viroli_dataset(
			n = n,
			means_list = means_components,
			U_list = U_list,
			V_list = V_list,
			seed = sim + 5000  # Different seed for each simulation
		)
		
		# Fit the model
		fit <- matrix_variate_mixture_fit(
			x_list = dataset$X,
			g = g,
			max_iter = 100,
			tol = 1e-6,
			verbose = FALSE
		)
		
		# Compute accuracy (accounting for label switching)
		pred_labels <- fit$cluster
		true_labels <- dataset$true_groups
		
		# Check both original and swapped label permutations
		# For 3 groups, we need to check all 3! = 6 permutations
		permutations <- list(
			c(1, 2, 3), c(1, 3, 2), c(2, 1, 3),
			c(2, 3, 1), c(3, 1, 2), c(3, 2, 1)
		)
		
		accuracies <- numeric(6)
		for (perm_idx in 1:6) {
			perm <- permutations[[perm_idx]]
			relabeled <- perm[pred_labels]
			accuracies[perm_idx] <- mean(relabeled == true_labels)
		}
		
		best_accuracy <- max(accuracies)
		best_perm <- permutations[[which.max(accuracies)]]
		aligned_labels <- best_perm[pred_labels]
		
		all_accuracies[sim] <- best_accuracy
		all_confusion_matrices[[sim]] <- table(True = true_labels, 
		                                        Predicted = aligned_labels)
		all_fits[[sim]] <- fit
	}
	
	list(
		accuracies = all_accuracies,
		mean_accuracy = mean(all_accuracies),
		sd_accuracy = sd(all_accuracies),
		min_accuracy = min(all_accuracies),
		max_accuracy = max(all_accuracies),
		confusion_matrices = all_confusion_matrices,
		fits = all_fits,
		means = means_components,
		U_covariances = U_list,
		V_covariances = V_list
	)
}

```

```{r}
#| label: viroli_run_simulation
#| warning: false
#| message: false

viroli_sim_results <- run_viroli_simulation(N_sim = 1, n = 300, g = 3, 
                                            verbose = TRUE)
```

```{r}
#| label: viroli_summary_table
#| warning: false
#| message: false

paste("Mean accuracy %:", viroli_sim_results$mean_accuracy)
paste("SD accuracy %:", viroli_sim_results$sd_accuracy)
paste("Min accuracy %:", viroli_sim_results$min_accuracy)
paste("Max accuracy %:", viroli_sim_results$max_accuracy)

summary_df <- data.frame(
	Simulation = 1:100,
	Accuracy = viroli_sim_results$accuracies)

print(head(summary_df[order(-summary_df$Accuracy), ], 10))
print(head(summary_df[order(summary_df$Accuracy), ], 10))
```

## In the style of Tomarchio

A total of 100 datasets are created with n = 200 $2 \times 4$ matrices in two groups of equal weight with parameters

$$
\begin{align}
M_1 = \begin{pmatrix}
        −2.60 & -1.10 & -0.50 & -0.20\\
        1.30 & 0.60 & 0.30 & 0.10
      \end{pmatrix}
\qquad
&M_2 = \begin{pmatrix}
        1.50 & 1.70 & 1.90 & 2.20\\
        -3.70 & -2.70 & -2.00 & -1.50
      \end{pmatrix} \\
      
U_1 = \begin{pmatrix}
        2 & 0 \\
        0 & 1
      \end{pmatrix}
\qquad
&U_2 = \begin{pmatrix}
        1.70 & 0.5 \\
        0.5 & 1.30
      \end{pmatrix}
\end{align}     
\\
      
V_1 = V_2 = \begin{pmatrix}
        1.00 & 0.50 & 0.25 & 0.13 \\
        0.50 & 1.00 & 0.50 & 0.25 \\
        0.25 & 0.50 & 1.00 & 0.50 \\
        0.13 & 0.25 & 0.50 & 1.00
      \end{pmatrix} \\
$$




```{r}
#| label: tomarchio_simulation_setup
#| warning: false
#| message: false

generate_tomarchio_means <- function() {
	M1 <- matrix(c(
		-2.60, -1.10, -0.50, -0.20,
		 1.30,  0.60,  0.30,  0.10
	), nrow = 2, byrow = TRUE)
	
	M2 <- matrix(c(
		 1.50,  1.70,  1.90,  2.20,
		-3.70, -2.70, -2.00, -1.50
	), nrow = 2, byrow = TRUE)
	list(M1 = M1, M2 = M2)
}

generate_tomarchio_covariances <- function() {
	U1 <- matrix(c(2.0, 0.0,
	               0.0, 1.0), nrow = 2, byrow = TRUE)
	U2 <- matrix(c(1.70, 0.5,
	               0.5, 1.30), nrow = 2, byrow = TRUE)
	V <- matrix(c(
		1.00, 0.50, 0.25, 0.13,
		0.50, 1.00, 0.50, 0.25,
		0.25, 0.50, 1.00, 0.50,
		0.13, 0.25, 0.50, 1.00
	), nrow = 4, byrow = TRUE)
	list(U1 = make_spd(U1), U2 = make_spd(U2), V = make_spd(V))
}

generate_tomarchio_dataset <- function(n = 200, means_list, cov_list, seed = NULL) {
	group_sizes <- c(floor(n / 2), n - floor(n / 2))
	group_labels <- rep(1:2, times = group_sizes)
	X_list <- vector("list", n)
	component_covs <- list(cov_list$U1, cov_list$U2)

	idx <- 1
	for (g in 1:2) {
		chol_U <- chol(component_covs[[g]])
		chol_V <- chol(cov_list$V)
		for (i in seq_len(group_sizes[g])) {
			Z <- matrix(rnorm(2 * 4), nrow = 2, ncol = 4)
			X_list[[idx]] <- means_list[[g]] + t(chol_U) %*% Z %*% chol_V
			idx <- idx + 1
		}
	}

	list(X = X_list, true_groups = group_labels)
}

run_tomarchio_simulation <- function(N_sim = 100, n = 200, verbose = TRUE) {
	means <- generate_tomarchio_means()
	cov_list <- generate_tomarchio_covariances()
	all_accuracies <- numeric(N_sim)
	all_fits <- vector("list", N_sim)

	for (sim in seq_len(N_sim)) {
		if (verbose && sim %% 10 == 0) {
			cat(sprintf("Tomarchio simulation %d/%d\n", sim, N_sim))
		}
		dataset <- generate_tomarchio_dataset(
			n = n,
			means_list = list(means$M1, means$M2),
			cov_list = cov_list,
			seed = 8000 + sim
		)
		fit <- matrix_variate_mixture_fit(
			x_list = dataset$X,
			g = 2,
			max_iter = 100,
			tol = 1e-6,
			verbose = FALSE
		)
		pred_labels <- fit$cluster
		true_labels <- dataset$true_groups
		acc1 <- mean(pred_labels == true_labels)
		acc2 <- mean(3 - pred_labels == true_labels)
		all_accuracies[sim] <- max(acc1, acc2)
		all_fits[[sim]] <- fit
	}

	list(
		accuracies = all_accuracies,
		mean_accuracy = mean(all_accuracies),
		sd_accuracy = sd(all_accuracies),
		min_accuracy = min(all_accuracies),
		max_accuracy = max(all_accuracies),
		means = means,
		covariances = cov_list,
		fits = all_fits
	)
}
```

```{r}
#| label: tomarchio_run_simulation
#| warning: false
#| message: false

tomarchio_sim_results <- run_tomarchio_simulation(N_sim = 100, n = 200, verbose = TRUE)
```


```{r}
#| label: tomarchio_results
#| warning: false
#| message: false

paste("Mean accuracy %:", tomarchio_sim_results$mean_accuracy)
paste("SD accuracy %:", tomarchio_sim_results$sd_accuracy)
paste("Min accuracy %:", tomarchio_sim_results$min_accuracy)
paste("Max accuracy %:", tomarchio_sim_results$max_accuracy)


summary_df <- data.frame(
	Simulation = 1:100,
	Accuracy = tomarchio_sim_results$accuracies
  )

print(head(summary_df[order(-summary_df$Accuracy), ], 10))
print(head(summary_df[order(summary_df$Accuracy), ], 10))
```

## Functional Data

Here we test the matrix-variate clustering model on functional data represented as image matrices (hearts vs random squares). Each observation is a 12×12 matrix with grayscale values and added noise.

```{r}
sim_function<-function(Ea,Eb,Ec,x,basis,fun){
  a=rnorm(30,Ea,0.4^2)
  b=rnorm(30,Eb,0.4^2)
  c=rnorm(30,Ec,0.4^2)
  y=a*fun(x-b)+c
  return(data.frame(basis,y=y))}

plotImage<-function(m){
  df <- as.data.frame(as.table(m))  
  df$Var1 <- as.numeric(df$Var1)
  df$Var2 <- as.numeric(df$Var2)
  
  ggplot(df, aes(x = Var2, y = Var1, fill = Freq)) +
    geom_tile() +
    scale_y_reverse() +       
    coord_fixed() +          
    scale_fill_gradient(low = "white", high = "black") +  
    theme_void()   +theme(legend.position = "none")
}

make_shape<-function(shape){
  if(shape=="heart"){
    m <- matrix(0, 12, 12)
    m[3, 4:5] <- 0.8
    m[3, 8:9] <- 0.8
    m[4, 3:6] <- 0.9
    m[4, 7:10] <- 0.9
    m[5, 2:11] <- 1.0
    m[6, 2:11] <- 1.0
    m[7, 3:10] <- 0.9
    m[8, 4:9] <- 0.8
    m[9, 5:8] <- 0.6
    m[10, 6:7] <- 0.4
  }else{
    m <- matrix(0, 12, 12)
    for (i in 4:9) {
      for (j in 4:9) {
        m[i, j] <- runif(1, 0.7, 1)  # random grayscale inside square
      }
    }}
  noise <- matrix(rnorm(length(m), mean = 0, sd = 0.1), nrow = nrow(m))
  noisy <- pmin(pmax(m + noise, 0), 1)  # clamp between 0 and 1
  return(noisy)
}
```


```{r}
#| label: functional_data
#| warning: false
#| message: false

set.seed(42)
n_per_group <- 200

# Heart shapes with noise
hearts <- replicate(n_per_group, make_shape("heart"), simplify = FALSE)

# Random squares with noise
squares <- replicate(n_per_group, make_shape("square"), simplify = FALSE)

functional_data <- c(hearts, squares)
true_labels_functional <- c(rep(1, n_per_group), rep(2, n_per_group))
```

```{r}
#| label: ex_func_clust
#| warning: false
#| message: false

fit_functional <- matrix_variate_mixture_fit(
	x_list = functional_data,
	g = 2,
	max_iter = 100,
	tol = 1e-6,
	nstart = 10,
	verbose = TRUE
)

pred_labels <- fit_functional$cluster
acc1 <- mean(pred_labels == true_labels_functional)
acc2 <- mean(3 - pred_labels == true_labels_functional)
accuracy <- max(acc1, acc2)

cat("Clustering Accuracy:", round(accuracy * 100, 2), "%\n")
# Iterations
fit_functional$iterations
# Final Loglikelihood
round(tail(fit_functional$logLik, 1), 4)

print(fit_functional$pi)
table(true = true_labels_functional, predicted = pred_labels)
```

```{r}
#| label: func_viz
#| warning: false
#| message: false
#| fig.width: 14
#| fig.height: 10

par(mfrow = c(4, 5), mar = c(2, 2, 2, 1))
for (i in 1:10) {
	image(functional_data[[i]], main = paste("Heart", i, "-> Cluster", pred_labels[i]), 
	      axes = FALSE)
}

for (i in (n_per_group + 1):(n_per_group + 10)) {
	image(functional_data[[i]], main = paste("Square", i - n_per_group, "-> Cluster", pred_labels[i]), 
	      axes = FALSE)
}
```

```{r}
#| label: func_diagnostics
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 5

plot(fit_functional$logLik, type = "l", xlab = "Iteration", ylab = "Log-Likelihood",
     main = "Convergence Trace: Functional Data Clustering",
     col = "steelblue", lwd = 2)
grid()

par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
for (g in 1:2) {
	image(fit_functional$M[[g]], main = paste("Estimated Mean - Component", g),
	      axes = FALSE, col = gray.colors(256))
}
par(mfrow = c(1, 1))
```
